Pigeons.jl: Distributed sampling from intractable distributions

### 2024-05-22 paper/PT.tex
Changes based on Trevor's comments
Update based on Trevor's comments

To resolve this issue, PT constructs a sequence of $N$ distributions,  
$\pi_1, \pi_2, \ldots, \pi_N$, where $\pi_N$ is usually equal to $\pi$.
The distributions are chosen so that it is easy to obtain samples from $\pi_1$
with the sampling difficulty increasing as one approches $\pi_N$. An example 
of such a sequence of distributions, referred to as an \textit{annealing path},
is shown in Figure~\ref{fig:path}.

@@

We now turn to explain how this path of distributions can be used to enhance 
sampling from the target distribution, $\pi$. 
PT operates by first obtaining samples from each distribution on the path in parallel 
(referred to as an \textit{exploration phase}). 
Then, samples between adjacent distributions are swapped (referred to as 
a \textit{communication phase}). The communication phase in PT is crucial: it allows 
for the discovery of new regions of the space of the target distribution such as 
the top-right mode of the distribution presented in Figure~\ref{fig:bimodal}.

@@

We study the weak scaling properties of Pigeons' implementation of PT
as given by Gustafson's law.
In this setup, we consider a varying problem size proportional to the number of processes used. 
The number of chains is our definition of the problem size; 
this is in line with previous heuristics that establish that the problem 
difficulty is generally proportional to the optimal number of chains in PT \cite{syed2021nrpt}.
In particular, we consider a multivariate normal target distribution 
and vary the dimension $d \in \cbra{2^1, 2^2, \ldots, 2^{12}}$ for the distributed setup and 
$d \in \cbra{2^1, 2^2, \ldots, 2^{10}}$ for the parallel (shared memory) setup. 
The number of chains is set to $\ceil{\sqrt{d}}$ according to PT tuning suggestions \cite{syed2021nrpt} 
and for the distributed and parallel settings we set the number of processes or threads equal 
to the number of chains.
Within each chain we use the autoMALA sampler \cite{biron2024automala}, which is a 
gradient-based sampler with automatic step size selection in the Metropolis-adjusted Langevin algorithm.
Ten different seeds are used in order to obtain approximate 95\% confidence intervals of 
the speedup.

@@

We emphasize again that Pigeons supports both shared memory 
parallelism (SMP) and distributed memory parallelism (DMP). However, in certain cases, it 
might be useful to run parallel tempering with very many chains or 
to run several instances of parallel tempering at once. In such cases, 
the DMP setup is the most natural if the requested number of chains is very large. 
Another notable example of the benefit of DMP is for the multi-GPU use case. 
For instance, in some situations the likelihood evaluation of a single chain 
requires its own GPU (e.g., for high-resolution black hole imaging, which relies 
on running a Fast Fourier Transform on a GPU). 
Additionally, DMP can be useful when the user's code is not thread safe. 
In such cases, concurrency can still be achieved via DMP.
Our implementation in Pigeons ensures that the 
communication cost in the DMP setup is minimal. In most cases, the 
communication cost is negligible compared to other computations performed on each process.

>>>>

\medskip
To resolve this issue, PT constructs a sequence of $N$ distributions,  
$\pi_1, \pi_2, \ldots, \pi_N$, where $\pi_N$ is usually equal to $\pi$.
The distributions are chosen so that it is easy to obtain samples from $\pi_1$
with the sampling difficulty increasing as one approches $\pi_N$. An example 
of such a sequence of distributions, referred to as an \textit{annealing path},
is shown in Figure~\ref{fig:path}.

@@

\medskip
We now turn to explain how this path of distributions can be used to enhance 
sampling from the target distribution, $\pi$. 
PT operates by first obtaining samples from each distribution on the path in parallel 
(referred to as an \textit{exploration phase}). 
Then, samples between adjacent distributions are swapped (referred to as 
a \textit{communication phase}). The communication phase in PT is crucial: it allows 
for the discovery of new regions of the space of the target distribution such as 
the top-right mode of the distribution presented in Figure~\ref{fig:bimodal}.

@@

We study the weak scaling properties of Pigeons' implementation of PT.
In this setup, we consider a varying problem size proportional to the number of processes used. 
The number of chains is our definition of the problem size; 
this is in line with previous heuristics that establish that the problem 
difficulty is generally proportional to the optimal number of chains in PT \cite{syed2021nrpt}.
In particular, we consider a multivariate normal target distribution 
and vary the dimension $d \in \cbra{2^1, 2^2, \ldots, 2^{12}}$ for the distributed setup and 
$d \in \cbra{2^1, 2^2, \ldots, 2^{10}}$ for the parallel (shared memory) setup. 
The number of chains is set to $\ceil{\sqrt{d}}$ according to PT tuning suggestions \cite{syed2021nrpt} 
and for the distributed and parallel settings we set the number of processes or threads equal 
to the number of chains.
Within each chain we use the autoMALA sampler \cite{biron2024automala}, which is a 
gradient-based sampler with automatic step size selection in the Metropolis-adjusted Langevin algorithm.
Ten different seeds are used in order to obtain approximate 95\% confidence intervals of 
the speedup.

@@

We emphasize again that Pigeons supports both shared memory 
parallelism (SMP) and distributed memory parallelism (DMP). However, in certain cases, it 
might be useful to run parallel tempering with very many chains or 
to run several instances of parallel tempering at once. In such cases, 
the DMP setup is the most natural if the requested number of chains is very large. 
Another notable example of the benefit of DMP is for the multi-GPU use case. 
For instance, in some situations the likelihood evaluation of a single chain may 
require its own GPU (e.g., if one is running a Fast Fourier Transform on a GPU). 
Additionally, DMP can be useful when the user's code is not thread safe. 
In such cases, concurrency can still be achieved via DMP.
Our implementation in Pigeons ensures that the 
communication cost in the DMP setup is minimal. In most cases, the 
communication cost is negligible compared to other computations performed on each process.

### 2024-05-22 paper/examples.tex
Update based on Trevor's comments

\medskip 
As a side-product of PT, we automatically obtain an approximation to the natural 
logarithm of the normalization constant. This is done automatically using the 
stepping stone estimator \cite{xie2011improving}.
The estimate can be accessed using
\begin{lstlisting}[language=Julia]
stepping_stone(pt)
\end{lstlisting}
In the case of the normalization constant given by \cref{eq:coinflip_normalization} 
for $n=100,000$ and $y=50,000$, we can exactly obtain its value as $\log(Z) \approx -11.8794$ 
using a computer algebra system. Note that this is very close to the output provided in 
\cref{fig:standard_out}.

>>>>

\medskip 
As a side-product of PT, we automatically obtain an approximation to the natural 
logarithm of the normalization constant. This is done automatically using the 
stepping stone estimator \cite{xie2011improving}.
The estimate can be accessed using
\begin{lstlisting}[language=Julia]
stepping_stone(pt)
\end{lstlisting}
In the case of the normalization constant given by \cref{eq:coinflip_normalization} 
for $n=100,000$ and $y=50,000$, we can exactly obtain its value as $\log(Z) \approx -11.8794$.
Note that this is very close to the output provided in \cref{fig:standard_out}.
